---
title: "STAT 639 - Homework 1"
subtitle: "ISLR Chapter 2: Exercises 1, 2, 8, and 10"
author: "Bao Dang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# Exercise 1

For each case, determine whether a more flexible statistical learning method would generally be expected to perform better or worse than a less flexible method.

## (a) Large number of observations, small number of predictors

A **more flexible method should generally perform better**.

When \(n\) is very large and \(p\) is small, there is enough data to estimate a flexible model without suffering too much from high variance. A flexible model can capture more complicated relationships between the predictors and response, so the reduction in bias may outweigh the increase in variance.

## (b) Small number of observations, large number of predictors

A **more flexible method should generally perform worse**.

With small \(n\) and large \(p\), a flexible method can easily fit random noise in the training data. This leads to overfitting and high variance. A less flexible method is generally safer in this setting.

## (c) Highly non-linear relationship between predictors and response

A **more flexible method should generally perform better**.

If the true relationship is highly non-linear, a flexible model is better able to approximate that relationship. A simple model may have high bias because it cannot capture the true structure.

## (d) Very high variance of the error terms

A **more flexible method should generally perform worse**.

When the irreducible noise is very large, a highly flexible method may begin fitting the noise rather than the underlying signal. This increases variance and can hurt test-set performance.

# Exercise 2

For each situation, identify whether the problem is classification or regression, whether the main goal is inference or prediction, and identify \(n\) and \(p\).

## (a) CEO salary

The response is CEO salary, and the explanatory variables include company profit, number of employees, and industry.

- **Type:** Regression
- **Goal:** Inference
- **\(n\):** 500 firms
- **\(p\):** 3 predictors

This is a regression problem because CEO salary is quantitative. The purpose is to understand which factors are associated with CEO salary, so the primary goal is inference.

## (b) Success of a new product

The response indicates whether a product is successful or unsuccessful. The predictors consist of price, marketing budget, competitors' price, and ten other variables.

- **Type:** Classification
- **Goal:** Prediction
- **\(n\):** 20 products
- **\(p\):** 13 predictors

This is classification because the response is qualitative. The goal is to predict whether the new product will succeed.

## (c) Weekly change in the U.S. dollar

The response is the weekly percentage change in the U.S. dollar relative to the euro, using weekly percentage changes in the stock market and the company's market as predictors.

- **Type:** Regression
- **Goal:** Prediction
- **\(n\):** 52 weeks
- **\(p\):** 2 predictors

The response is quantitative, so this is regression. Since the goal is to forecast the weekly percentage change, the main goal is prediction.

# Exercise 8 - College Data

This exercise uses the `College` data set described in ISLR Chapter 2.

## (a) Read the data

The code below first looks for `College.csv` in the current working directory. If it is not present, it downloads the file from the book website.

```{r ex8-load}

college_file <- "College.csv"

if (!file.exists(college_file)) {
  download.file(
    "https://www.statlearning.com/s/College.csv",
    destfile = college_file,
    mode = "wb"
  )
}

college <- read.csv("College.csv", check.names = FALSE)

```

There are `r nrow(college)` colleges and universities. Before removing the institution-name column, the imported data contain `r ncol(college)` columns.

## (b) Use college names as row names

```{r ex8-row-names}

# Inspect first column
head(college[[1]])
anyDuplicated(college[[1]])

# Only use it as row names if it is unique
if (anyDuplicated(college[[1]]) == 0) {
  rownames(college) <- college[[1]]
  college <- college[, -1, drop = FALSE]
}else{
  rownames(college) <- college[, 1]
  college <- college[, -1]
}

dim(college)
head(college)
```

After removing the institution-name column, the data contain 777 observations and 18 variables. The college names are now stored as row names rather than as a predictor.