---
title: "STAT 639 - Homework 1"
subtitle: "ISLR Chapter 2: Exercises 1, 2, 8, and 10"
author: "Bao Dang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# Exercise 1

For each case, determine whether a more flexible statistical learning method would generally be expected to perform better or worse than a less flexible method.

## (a) Large number of observations, small number of predictors

A **more flexible method should generally perform better**.

When \(n\) is very large and \(p\) is small, there is enough data to estimate a flexible model without suffering too much from high variance. A flexible model can capture more complicated relationships between the predictors and response, so the reduction in bias may outweigh the increase in variance.

## (b) Small number of observations, large number of predictors

A **more flexible method should generally perform worse**.

With small \(n\) and large \(p\), a flexible method can easily fit random noise in the training data. This leads to overfitting and high variance. A less flexible method is generally safer in this setting.

## (c) Highly non-linear relationship between predictors and response

A **more flexible method should generally perform better**.

If the true relationship is highly non-linear, a flexible model is better able to approximate that relationship. A simple model may have high bias because it cannot capture the true structure.

## (d) Very high variance of the error terms

A **more flexible method should generally perform worse**.

When the irreducible noise is very large, a highly flexible method may begin fitting the noise rather than the underlying signal. This increases variance and can hurt test-set performance.

# Exercise 2

For each situation, identify whether the problem is classification or regression, whether the main goal is inference or prediction, and identify \(n\) and \(p\).

## (a) CEO salary

The response is CEO salary, and the explanatory variables include company profit, number of employees, and industry.

- **Type:** Regression
- **Goal:** Inference
- **\(n\):** 500 firms
- **\(p\):** 3 predictors

This is a regression problem because CEO salary is quantitative. The purpose is to understand which factors are associated with CEO salary, so the primary goal is inference.

## (b) Success of a new product

The response indicates whether a product is successful or unsuccessful. The predictors consist of price, marketing budget, competitors' price, and ten other variables.

- **Type:** Classification
- **Goal:** Prediction
- **\(n\):** 20 products
- **\(p\):** 13 predictors

This is classification because the response is qualitative. The goal is to predict whether the new product will succeed.

## (c) Weekly change in the U.S. dollar

The response is the weekly percentage change in the U.S. dollar relative to the euro, using weekly percentage changes in the stock market and the company's market as predictors.

- **Type:** Regression
- **Goal:** Prediction
- **\(n\):** 52 weeks
- **\(p\):** 2 predictors

The response is quantitative, so this is regression. Since the goal is to forecast the weekly percentage change, the main goal is prediction.

# Exercise 8 - College Data

This exercise uses the `College` data set described in ISLR Chapter 2.

## (a) Read the data

The code below first looks for `College.csv` in the current working directory. If it is not present, it downloads the file from the book website.

```{r ex8-load}

college_file <- "College.csv"

if (!file.exists(college_file)) {
  download.file(
    "https://www.statlearning.com/s/College.csv",
    destfile = college_file,
    mode = "wb"
  )
}

college <- read.csv("College.csv", check.names = FALSE)

```

There are `r nrow(college)` colleges and universities. Before removing the institution-name column, the imported data contain `r ncol(college)` columns.

## (b) Use college names as row names

```{r ex8-row-names}

# Inspect first column
rownames(college) <- college[, 1]
View(college)
```

After removing the institution-name column, the data contain `r nrow(college)` observations and `r ncol(college)` variables. The college names are now stored as row names rather than as a predictor.

## (c-i) Numerical summary

```{r ex8-summary}
summary(college)
```

The variables differ substantially in scale. For example, enrollment and application counts vary widely across schools, as do tuition and instructional expenditures. Graduation rates and percentages such as `Top10perc`, `PhD`, and `perc.alumni` are naturally bounded percentage variables.

## (c-ii) Scatterplot matrix of the first ten variables

```{r ex8-pairs}

# check first 10 columns
str(college[, 1:10])

rownames(college) <- college[, 1]
college <- college[, -1]

summary(college)

# convert Private to factor before processing
college$Private <- factor(college$Private)

pairs(
  college[, 1:10],
  main = "Scatterplot Matrix: First 10 College Variables"
)
```

Several size-related variables are strongly associated. In particular, colleges receiving more applications also tend to accept and enroll more students. Full-time undergraduate enrollment is also associated with these measures of school size.

## (c-iii) Out-of-state tuition versus private/public status

```{r ex8-outstate-private}

plot(
  college$Private,
  college$Outstate,
  xlab = "Private",
  ylab = "Out-of-State Tuition",
  main = "Out-of-State Tuition by Private/Public Status"
)

```

Private colleges generally have substantially higher out-of-state tuition than public colleges.

## (c-iv) Create an Elite variable

A school is classified as elite when more than 50% of its students come from the top 10% of their high-school class.

```{r ex8-elite}
Elite <- rep("No", nrow(college))
Elite[college$Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)

college <- data.frame(college, Elite)

summary(college$Elite)
```

The number of elite colleges is:
  
```{r ex8-elite-count}
sum(college$Elite == "Yes")
```

For the standard ISLR `College` data, this gives **`r sum(college$Elite == "Yes")` elite colleges**.

Now compare tuition between elite and non-elite schools.

```{r ex8-outstate-elite}
plot(
  college$Elite,
  college$Outstate,
  xlab = "Elite",
  ylab = "Out-of-State Tuition",
  main = "Out-of-State Tuition by Elite Status"
)
```

Elite colleges tend to have higher out-of-state tuition than non-elite colleges.

## (c-v) Histograms

```{r ex8-histograms}
par(mfrow = c(2, 2))

hist(
  college$Apps,
  breaks = 20,
  main = "Applications",
  xlab = "Apps"
)

hist(
  college$Outstate,
  breaks = 20,
  main = "Out-of-State Tuition",
  xlab = "Outstate"
)

hist(
  college$Expend,
  breaks = 20,
  main = "Instructional Expenditure",
  xlab = "Expend"
)

hist(
  college$Grad.Rate,
  breaks = 20,
  main = "Graduation Rate",
  xlab = "Grad.Rate"
)

par(mfrow = c(1, 1))
```

Applications and instructional expenditure show noticeable right-skewness, reflecting a relatively small number of very large or high-spending institutions. Graduation rates are more concentrated toward the upper part of their possible range.

## (c-vi) Additional exploration

```{r ex8-extra}
# Correlations among selected quantitative variables
selected <- college[, c(
  "Apps", "Accept", "Enroll", "Top10perc",
  "Outstate", "Room.Board", "PhD",
  "Expend", "Grad.Rate"
)]

round(cor(selected), 2)

# Graduation rate versus instructional expenditure
plot(
  college$Expend,
  college$Grad.Rate,
  xlab = "Instructional Expenditure per Student",
  ylab = "Graduation Rate",
  main = "Graduation Rate vs. Expenditure"
)

# Top 10% students versus tuition
plot(
  college$Top10perc,
  college$Outstate,
  xlab = "% of Students from Top 10% of High School Class",
  ylab = "Out-of-State Tuition",
  main = "Selectivity vs. Tuition"
)
```

Overall, the data suggest strong relationships among measures of institution size. More selective colleges also tend to charge higher tuition, and private or elite colleges generally have higher out-of-state tuition.

# Exercise 10 - Boston Housing Data

## Setup

```{r ex10-setup}
if (!requireNamespace("ISLR2", quietly = TRUE)) {
  install.packages("ISLR2")
}

library(ISLR2)
data("Boston")
```

## (a) Dimensions and meaning of rows and columns

```{r ex10-dimensions}
?Boston

dim(Boston)
names(Boston)
str(Boston)
```

The Boston data contain **`r nrow(Boston)` rows and `r ncol(Boston)` columns**.

Each row represents a Boston-area census tract. The columns contain tract-level predictors such as crime rate, zoning, industrial land use, air pollution, average number of rooms, property tax rate, pupil-teacher ratio, and related variables. The `medv` variable is the median value of owner-occupied homes.

## (b) Pairwise scatterplots

```{r ex10-pairs}
pairs(
  Boston,
  main = "Boston Housing Data: Pairwise Scatterplots"
)
```

Some clear relationships are visible. For example, `medv` tends to increase with `rm` and decrease with `lstat`. Crime tends to be higher in areas with higher `rad` and `tax`, and lower in areas with larger `dis`.

## (c) Predictors associated with per-capita crime rate

```{r ex10-crime}
crime_cor <- sort(
  cor(Boston)[, "crim"],
  decreasing = TRUE
)

crime_cor
```

```{r ex10-crime-plots}
par(mfrow = c(2, 2))

plot(Boston$rad, Boston$crim,
     xlab = "Accessibility to Radial Highways (rad)",
     ylab = "Crime Rate (crim)")

plot(Boston$tax, Boston$crim,
     xlab = "Property Tax Rate (tax)",
     ylab = "Crime Rate (crim)")

plot(Boston$dis, Boston$crim,
     xlab = "Distance to Employment Centers (dis)",
     ylab = "Crime Rate (crim)")

plot(Boston$lstat, Boston$crim,
     xlab = "Lower-Status Population % (lstat)",
     ylab = "Crime Rate (crim)")

par(mfrow = c(1, 1))
```

Crime rate is positively associated with variables such as `rad`, `tax`, `indus`, `nox`, and `lstat`, and negatively associated with variables such as `dis`, `rm`, and `medv`. These are associations only and should not be interpreted as causal effects.




